\chapter{Introduction}

\section{Thesis}

\emph{It is practical and useful to design and implement an optional typing system 
for the Clojure programming language using bidirectional checking that allows Clojure programmers to continue 
using idioms and style found in current Clojure code.}

\section{Motivation}

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This idea is not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
are noteworthy for their broad success in matching many of the advantages of statically typed
languages (see work by Tobin-Hochstadt~\cite{Tob10}), 
notably due to the use of bidirectional checking (like work by Pierce and Turner~\cite{PT00}).
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

The Clojure programming language is a dynamically typed dialect of Lisp invented
by Hickey~\cite{Hic08}, designed to run on popular platforms.
It emphasises functional programming with immutable data structures
and provides direct interoperability with its host platform.
Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
the Common Language Runtime, and for Javascript virtual machines.
At the current time, Clojure on the JVM is the most mature implementation,
and therefore this project focuses on the JVM implementation.

Clojure has attracted wide-spread users in part by concentrating on pragmatism.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.
Also, Clojure's extensive host interoperability offers Clojure programmers
access to existing libraries for their platform, such as the vast Java library ecosystem.
By coupling pragmatic necessities with elegant features like Lisp-style macros, functional programming,
and immutability by default, Clojure is a compelling general purpose programming language.

Recently a number of languages have been created or modified to support aspects
of both static and dynamic typing.
Dart~\cite{Dart2012} (created by Google) is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript~\cite{TypeS2012} (created by Microsoft) adds an optional type system to Javascript,
a well-known dynamic language.
Typed Racket~\cite{TF08,Tob10} (by Tobin-Hochstadt et al.) goes further by offering safe interoperability
between typed and untyped modules by generating appropriate runtime assertions based
on expected static types.

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. For example, ``design by contract'' is often used,
introduced by Meyer for the Eiffel language~\cite{Mey92},
in which the programmer defines contracts that are enforced at runtime.
Unit testing is also a popular verification technique in dynamic languages.
Clojure adopts these approaches, providing easy syntax for defining 
Eiffel-inspired pre- and post-conditions and a library for writing unit tests.

Static type systems, however, are still desirable for many situations.
Powerful type systems like ML's~\cite{Mil97} and Haskell's~\cite{Mar10} have proved particularly useful
when complicated programming styles are required. For example,
Haskell's advanced static type system helps the programmer write correct monadic code
(as detailed by Wadler~\cite{Wad95})
especially in more complicated situations like combining monads via monad transformers.

\subsection{Why implement an optional type system for Clojure?}

\subsubsection{Better error messages}

Dynamically typed languages like Clojure often sacrifice some of the advantages of
static typing. In particular, statically typed languages generally offer 
compile-time type errors where dynamically typed languages either give runtime
type errors, or none at all.

This is distinction is easily demonstrated by comparing Clojure and Haskell's behaviour at their interactive prompts.
Listing \ref{lst:cljrtime} shows how Clojure happily compiles a nonsensical function that performs addition
on a string. It is only until we invoke it that we get a type error. 

\begin{lstlisting}[caption=Runtime type errors in Clojure, label=lst:cljrtime]
user=> (fn [x] (+ "string" x))
#<user$eval128376$fn__128377 user$eval128376$fn__128377@183ab4e>

user=> ((fn [x] (+ "string" x)) 1)
#<ClassCastException java.lang.ClassCastException: java.lang.String cannot be cast to java.lang.Number>
\end{lstlisting}

The error provided by Clojure is unsatisfying:

\begin{itemize}
  \item The error message does little to help the programmer identify the source of the error.
  \item It requires at least some knowledge of Clojure's internals to debug.
  \item We would rather receive the error at the earliest time possible: compile-time.
\end{itemize}

Listing \ref{lst:hskctime} shows how Haskell, a statically typed language, handles this interaction.
Haskell detects our programming error at compile-time, disallowing our function to compile.

\begin{lstlisting}[caption=Compile-time type errors in Haskell, label=lst:hskctime]
Prelude> (\x -> "string" + x)

<interactive>:1:17:
   No instance for (Num [Char])
   arising from a use of `+'
   Possible fix: add an instance declaration for (Num [Char])
   In the expression: "string" + x
   In the expression: (\ x -> "string" + x)
   In an equation for `it': it = (\ x -> "string" + x)
\end{lstlisting}

Haskell's error here is much preferred over Clojure's:

\begin{itemize}
  \item The exact source of the error is given in \emph{user code}.
  \item The error is given as early as possible: compile-time.
\end{itemize}

By adding static type checking to Clojure, we can achieve errors messages that are closer to
Haskell's.
Listing \ref{lst:tcctime} gives an idea of what kind of error messages Typed Clojure actually
provides in these situations. 
However, it deserves some explanation:

\begin{itemize}
  \item The given source of the error may be surprising, but it is explained
        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
        \lstinline|(. clojure.lang.Numbers add "string" a)|.
  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
  \item \lstinline|cf| type checks a form.
  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
\end{itemize}

\begin{lstlisting}[caption=Compile-time errors in Typed Clojure, label=lst:tcctime]
user=> (cf (fn> [[a :- Number]] (+ "string" a)))
#<Exception java.lang.Exception: 5: 
Static method clojure.lang.Numbers/add could not be applied to arguments:
Domains: 
  typed.core/AnyInteger typed.core/AnyInteger
  java.lang.Number java.lang.Number

Arguments:
  (Value "string") java.lang.Number

in: (. clojure.lang.Numbers add "string" a)>
\end{lstlisting}

Typed Clojure lists each of the expected domain types, of which the static method \lstinline|add| of the Java class
\lstinline|clojure.lang.Numbers| has been assigned two,
and also lists the types of the given arguments.
A line number and the source of the error is also provided.

Most importantly: the error was caught at \emph{compile-time}.

\subsubsection{Higher order programming}

The initial motivation for implementing an optional type system
for Clojure was outlined in a discussion with Jim Duey, a Clojure programmer, at Clojure Conj 2011.
In an apparently heroic effort, Duey managed to
implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
an advanced form of ``pipes'', using arrows, a generalisation of monads.
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

Duey highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness of the library.
Without a static type system, it is a significant task
to verify such an implementation as correct due to heavy use
of higher-order functions.
Secondly, to aid him while writing the library.

Programmers in statically typed languages take advantage of the static type system
to help verify code written in abstract programming styles, like monads.
Adding a static type system to Clojure should bring similar benefits to Clojure programmers.

\subsection{Why does Clojure not already feature static type checking?}

Given the advantages of static type checking outlined in the previous section, it is natural to ask:
why does Clojure not already support static type checking?
The simplest answer is that Clojure was not designed to support a static type system.
Instead, features and idioms that are not necessarily compatible with traditional type systems
were favoured, like variable-arity functions, untyped hash-maps used like objects, and dynamic extension of program behaviour.

Without specifically speaking for the designers of Clojure, at the time Clojure was designed
it was not clear how to reconcile Clojure's feature-set with static type checking in a satisfying way.
Since then, Typed Racket, which add static type checking to the Racket programming language,
has matured and gained acceptance.
Once this occurred, it seemed only a matter of time until the ideas developed for the Typed Racket project 
were tested for their applicability to Clojure. 
As far as we know, this project is the first serious attempt to build an optional static type system for Clojure.

\subsection{What kind of type system does Typed Clojure provide?}

% Page 1 Rowan's PhD

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey02},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.
This distinction has a long history, originating in work in the $\lambda$-calculus 
by Church~\cite{Chu40} and Curry~\cite{Cur34},
leading to intrinsic types and extrinsic types also being called Church types and
Curry types often in the literature.

An ordinary static type system is used to check whether programs are basically
meaningful. Pfenning and Reynolds call these type systems \emph{intrinsic}. A language with an intrinsic
static type system has a run-time semantics that depends on the types associated
with variables and expressions during type checking.
For example, C, Java, ML, and Haskell have intrinsic types.
This means programs written in these languages must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
does not depend on a static type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exactly one type,
a view advocated by Harper~\cite{Har12}
and common in the literature on static types (for example, Pierce~\cite{Pie02}).

\subsection{What is the community reaction to Typed Clojure?}

Since I began developing Typed Clojure, the Clojure community has shown significant interest in this work:


\begin{itemize}
  \item I developed this project as a Google Summer of Code 2012 project,
        after it was selected by the Clojure community as a Clojure Google Summer of Code
        project for 2012~\cite{GSOC2012}.
  \item I gave a talk on this project at the Clojure Conj
        2012 conference, the main international programming conference related to Clojure, in November.
\end{itemize}

\section{Contributions}

In this dissertation we develop an optional type system for Clojure.
We focus on the features needed to make a practical type system that
existing Clojure programmers can use.
We also develop a prototype based on Typed Racket~\cite{Tob10}.

The contributions of the work described in this dissertation are as follows:

\begin{itemize}
  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
        which is able to type check many common Clojure idioms.
        We show how this type checker can be used to aid Clojure programmers use sophisticated
        programming styles by porting most of an existing library for monadic programming.

  \item This prototype is extended to support several features that are not yet present in Typed Racket.
        We support filtering sequences on anonymous functions that convey only negative type information
        (eg. non-nil), capturing a common Clojure idiom.
        We enable type checking of very abstract code (like code using monads)
        by adding \emph{type functions} (functions at the type level).

  \item We support safer interoperability with Java from Clojure. We show how techniques developed for
        Typed Racket, such as occurrence typing~\cite{TF10}, help detect misuses of Java's \emph{null}
        while still conforming to existing Clojure idioms.

  \item We can type check several particularly useful Clojure idioms, like common usages of the \emph{sequence abstraction}
        (Clojure's common interface to all collections) and 
        common usages of ``untyped'' hash-maps, which have a similar role to records in ML or Haskell,
        and objects in object-oriented languages.

  \item We identify the main future issues in typing Clojure code. Multimethod definitions are problematic
        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
        Records have subtle semantics for dissociating keys, specifically: dissociating all base keys of a record returns
        a plain map.
\end{itemize}

Together these contributions support my thesis and motivate further work in this area.

\section{Typed Clojure through Examples}

% Hello world
This section introduces Typed Clojure with example code. 
Typed Clojure is developed for the JVM implementation of Clojure, therefore
the rest of this chapter uses that implementation.
An attempt is
made to introduce some Clojure syntax and semantics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is handy, but a brief tutorial is given
for newcomers.

\subsection{Preliminary: Lisp Syntax}

The core of understanding Lisp syntax when coming from a popular language
like Java or Javascript can be summarised by these points.

\begin{itemize}
  \item Operators are always in prefix position.
  \item Invocations are always wrapped in a pair of balanced parenthesis.
  \item Parenthesis start to the left of the operator.
\end{itemize}

For example, the Java expressions \emph{(1 + 2) / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
and \emph{numberCrunch(1, 2)} written \lstinline|(numberCrunch 1 2)|.

Clojure also adds other syntax:

\begin{itemize}
  \item Prefixing \lstinline|:| to a symbol defines a \emph{keyword}, often used for map keys. eg. \lstinline|:my-keyword|.
  \item Square brackets delimit vector literals. eg. \lstinline|[1 2]| is a 2 place vector.
  \item Curly brackets define map literals. eg. \lstinline|{:a 1 :b 2}| is a map from 
        \lstinline|:a| to \lstinline|1| and \lstinline|:b| to \lstinline|2|.
  \item Commas are always optional and treated as whitespace, but often used to show the intended structure.
\end{itemize}

\subsection{Simple Examples}

We begin with the obligatory \emph{Hello world} example.

\begin{lstlisting}[caption=Typed Hello world, label=lst:helloworld]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]))

(println "Hello world")
\end{lstlisting}

At this point, it is worth understanding Clojure's namespacing feature.
Clojure code is always executed in a \emph{namespace}, and each file of Clojure code should 
have a \lstinline|ns| declaration with the namespace name and its dependencies,
which switches the current namespace and executes the given dependency commands.
There is one special namespace, \lstinline|clojure.core| which is
loaded with every namespace, implicitly ``referring'' all its vars in the namespace.
For example, \lstinline|ns| refers to the var \lstinline|clojure.core/ns|,
similarly \lstinline|println| refers to \lstinline|clojure.core/println|
(vars are global bindings in Clojure).

The example in listing \ref{lst:helloworld} declares a dependency to 
\lstinline|typed.core|, Typed Clojure's main namespace. It also refers the var \lstinline|typed.core/check-ns|
into scope (\lstinline|check-ns| is the top level function for type checking a namespace).
Other than this dependency, this is identical to the untyped \emph{Hello world}.

More complex code may require extra annotations to type check:

\begin{lstlisting}[caption=Annotating vars in Typed Clojure (adapted from a Typed Scheme/Racket example by Tobin-Hochstadt~\cite{Tob10})]
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}

(\lstinline|(defn collatz [n] ...)| expands to \lstinline|(def collatz (fn [n] ...))|,
where \lstinline|(def name init)| defines a new var in the current namespace 
called \lstinline|name|, and \lstinline|fn| creates a function value).

In this example, we define a new var \lstinline|collatz|
(when unambiguous, I omit the qualifying namespace/package for the remainder of the chapter).
Typed Clojure type checks at the namespace granularity. All vars that occur
in a ``typed'' namespace must have a type annotation. 
Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
annotates \lstinline|collatz| to be a function from \lstinline|Number| to 
\lstinline|Number| (\lstinline|Number| refers to \lstinline|java.lang.Number|,
due to every Clojure namespace implicitly importing all Classes in the \lstinline|java.lang| package,
the equivalent of Java's \lstinline|import java.lang.*;|).

\subsection{Datatypes and Protocols}
\label{sec:datatypes}

As well as \lstinline|def| and \lstinline|defn| definitions,
Clojure programmers typically include datatype and protocol
definitions. 
Protocols are similar to Java interfaces~\cite{InterfaceTutorial} and datatypes are similar to classes implementing interfaces.
We can annotate datatype and protocol definitions similarly.
%bump over the page
\\\\\\\\\\
\begin{lstlisting}[caption=Annotating protocols and datatypes in Typed Clojure]
(ns typed.test.deftype
  (:require 
    [typed.core :refer [check-ns ann-datatype defprotocol>
                        ann-protocol AnyInteger]]))

(ann-protocol Age 
  age [Age -> AnyInteger])
(defprotocol> Age
  (age [this]))

(ann-datatype Person 
  [name :- String
   age :- AnyInteger])
(deftype Person [name age]
  Age
  (age [this] age))

(age (Person. "Lucy" 34))
\end{lstlisting}

\lstinline|defprotocol>| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
with a set of methods. It is identical to the usual \lstinline|defprotocol|, but is required when using Typed Clojure
due to its complicated macroexpansion.
\lstinline|ann-protocol| is a Typed Clojure macro for annotating 
a protocol with the types of its methods.
In this example, we define a protocol \lstinline|Age| with an \lstinline|age| method,
which is really a first-class function taking the target object as the first parameter. The
type signature provided with \lstinline|ann-protocol|, here \lstinline|[Age -> AnyInteger]|
which is the type of the function taking an \lstinline|Age| and returning an \lstinline|AnyInteger|,
is for this function.

\lstinline|deftype|
defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes.}
in the current namespace with a number of fields and methods. 
\lstinline|typed.core/ann-datatype| is a Typed Clojure form for annotating datatypes, including its field types.
In this example, we create a datatype \lstinline|typed.test.person.Person|
(datatype defintions generate a Java Class, where the current namespace is used as a starting point for its
qualifying package)
with fields \lstinline|name| and \lstinline|age| and 
extend the \lstinline|Age| protocol by implementing the \lstinline|age| method.

Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
in the order they are passed to \lstinline|deftype|
\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
instance, setting the fields to their corresponding positional arguments.
Typed Clojure checks the datatype constructor to be the equivalent of 
\lstinline|[String AnyInteger -> Person]|.

Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.
Since generally we only need to annotate definitions and not uses, the number of annotations is relatively small.

\subsection{Polymorphism}

Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
can be restricted by \emph{bounds}.
In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
variable being bounded. Typed Clojure supports upper and lower type variable bounds.
This would be required mainly for future work involving Java Generics, like typing
invocations of \emph{java.util.Collection/max}\footnote{\url{http://docs.oracle.com/javase/1.5.0/docs/api/java/util/Collections.html#max(java.util.Collection)}}.

For accurate type checking, Typed Clojure parameterises some of Clojure's data structures. For example,
the interface behind Clojure's \emph{seq} abstraction for sequences \lstinline|clojure.lang.Seqable| has one 
covariant parameter\footnote{Covariant means that \lstinline|(Seqable Integer)| is a subtype of \lstinline|(Seqable Number)|
because Integer is a subtype of Number.}.

\begin{lstlisting}[caption=Polymorphism in Typed Clojure]
...
(ann to-set 
     (All [x]
       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
(defn to-set [a]
  (set a))
...
\end{lstlisting}

In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}, 
we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
\lstinline|All| introduces a set of type variables to the body of a type,
here \lstinline|x| is used to define a relationship between the input type and return type.

\lstinline|(U nil (Seqable x))| is a common type in Typed Clojure, read as the union
of the type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
it as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.
The ability to specify types that explicitly include or exclude \lstinline|nil| is one of the strengths
of Typed Clojure, and an aspect where it is more expressive than standard type systems like that for Java.

\subsection{Singleton Types}

Following Typed Racket, singleton types for certain values are provided
in Typed Clojure.
A singleton type is a type with a single member, like \lstinline|1|,
\lstinline|:a|, or \lstinline|"a"|\footnote{Strings are delimited by " in Clojure}.
Typed Clojure provides syntax for singleton types, either by passing
the value literal to the \lstinline|Value| primitive, or by prefixing
a quote (').

\begin{lstlisting}[caption=Singleton Types, label=lst:singletoneeg]
(ann k ':my-keyword)
(def k :my-keyword)
\end{lstlisting}

Listing \ref{lst:singletoneeg} shows a simple example of using
singleton types in Typed Clojure.
Singleton types are discussed further in section \ref{sec:intersectionunion}.

\subsection{Heterogeneous Maps}

It is idiomatic in Clojure to use plain maps with known keyword keys as lightweight objects.
Clojure provides a hash-map literal (using curly braces), making this idiom particularly
convenient.
For example,
\lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|
to value \lstinline|1|, and keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included occasionally for readability.

Typed Clojure provides heterogeneous map types which captures this common
``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
information on the types of key-value entries. In other words, it conveys
whether a particular key is present, but not whether it is absent.
The implications of this is discussed in section \ref{ref:designhmap}.
Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
in the syntax for defining heterogeneous map types.
%bump
\\\\
\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
(ann config '{:file String, :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
\end{lstlisting}

This example checks \lstinline|config| to be a heterogeneous map
with \lstinline|:file| and \lstinline|:ns| keys, with values of
type \lstinline|String| and \lstinline|Symbol| respectively.

Heterogeneous vector and seq types are also provided and work similarly,
except for their lack of keys like maps, hence are just a sequence of types.

\subsection{Variable-Arity Functions}

Functions in Clojure are multi-arity (a very commonly used feature) which means a function
can be defined with several sets of function arguments and bodies (or \emph{arities})
and which arity is executed depends on the number of arguments passed
to the function. A function can have any number of arities with fixed parameters, and at most one arity with
variable-parameters. Each arity must have a unique number of
parameters and have a lower number of parameters than the ``variable arity'', if present.

Strickland, Tobin-Hochstadt, and Felleisen invented a calculus and corresponding
implementation in Typed Racket for variable-arity polymorphism~\cite{STF09}
that is sufficient to handle uniform and non-uniform variable-arity functions.
Typed Clojure includes an implementation of the most immediately useful parts of variable-arity
polymorphism using algorithms, nomenclature, and implementation based on this work.

\subsubsection{Uniform Variable-Arity Functions}

A function with uniform variable parameters can treat its variable parameter
as a homogeneous list. 
Strickland et al.\ attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
in a function type, where \emph{T} is some type, and we take an identical approach.
For example, \lstinline|+| in Clojure accepts any number of arguments
of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.
\\\\\\\\

%bump over page
\begin{lstlisting}[caption=Typing multi-arity functions, label=lst:noteq]
(ann not= (Fn [Any -> boolean]
              [Any Any -> boolean]
              [Any Any Any * -> boolean]))
(defn not=
  "Same as (not (= obj1 obj2))"
  ([x] false)
  ([x y] (not (= x y)))
  ([x y & more]
   (not (apply = x y more))))
\end{lstlisting}

It is common to find Clojure library functions that define seemingly redundant
function arities for performance reasons.
Listing \ref{lst:noteq} defines the multi-arity function \lstinline|not=|,
taken from the Clojure standard library \lstinline|clojure.core| that uses this pattern.
\lstinline|not=| has three arities in its definition, including
one that takes a variable number of arguments.
The \lstinline|Fn| type constructor builds an \emph{ordered function intersection type} from
function types. Each arity must have at least one matching function type associated with it.
The two ``fixed arities'' are given familiar function types, with one and two fixed parameters respectively.
The type given for the arity with variable arguments \lstinline|[Any Any Any * -> boolean]|
uses a starred pre-type to signify any number of arguments of type \lstinline|Any|
can be provided to the right of its fixed arguments.

\subsubsection{Non-uniform Variable-Arity Functions}

Where \emph{uniform} variable-arity function types use \emph{starred pre-types}, \emph{non-uniform}
variable-arity function types use \emph{dotted pre-types}.
Typed Clojure supports usages of \emph{non-uniform} variable-arity functions,
where the variable parameter is a heterogeneous list, represented by a \emph{dotted type variable}.

For example, the variable argument function \lstinline|clojure.core/map| takes a function and one or more sequences,
and returns a sequence of the results of applying the function to each element of the sequences simultaneously.
Its type is given in listing \ref{lst:maptype}.

\begin{lstlisting}[caption=Type signature for \lstinline|clojure.core/map|, label=lst:maptype]
(ann clojure.core/map
     (All [c a b ...]
       [[a b ... b -> c] (U nil (Seqable a)) (U nil (Seqable b)) ... b -> (LazySeq c)]))
\end{lstlisting}

By adding \lstinline|...| after the last type variable in an \lstinline|All| binder
we can introduce a \emph{dotted type variable} into scope, which is a placeholder for a sequence of types.
A \emph{dotted pre-type} \lstinline|T ... b| over \emph{base} \lstinline|T| (a type) and \emph{bound}
\lstinline|b| (a dotted type variable)
serves as a placeholder for this sequence of types.
Dotted pre-types must appear to the right of all fixed parameters in a function type,
and cannot be mixed with other kinds of variable parameters like starred pre-types.
When a sequence of types of length \emph{n} is associated with a dotted pre-type, 
the dotted pre-type is expanded to \emph{n} copies of \lstinline|T|.
One other special property of a dotted pre-type is that the bound \lstinline|b|
is \emph{in scope} as a \emph{normal} type variable in its base \lstinline|T|.

To demonstrate dotted pre-types, we use \lstinline|typed.core/inst| to instantiate
\lstinline|map|. \lstinline|inst| takes a polymorphic expression and a number of types
that are satisfactory for instantiating the polymorphic type and returns an expression
of the instantiated polymorphic type.

The instantiation

\begin{lstlisting}
(inst map Number boolean String)
\end{lstlisting}

returns an expression of type

\begin{lstlisting}
[[boolean String -> Number] (U (Seqable boolean) nil) 
  (U (Seqable String) nil) -> (LazySeq Number)]
\end{lstlisting}

However, if sufficient types are given, the instantiation for \lstinline|map| can be inferred.
The invocation

\begin{lstlisting}
(map (ann-form (fn [a b c] c) 
               [boolean String Number -> Number]) 
     [true false] ["mystr" "astr"] [1 4])
\end{lstlisting}

uses \lstinline|typed.core/ann-form| to assign an expected type to
the first argument,
which is sufficient to infer the result type

\begin{lstlisting}
(LazySeq Number)
\end{lstlisting}


\subsection{Occurrence Typing}

It is common in Clojure, like other dynamically typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2}, conditional tests are used
to refine the types of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing~\cite{TF10} is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in Section \ref{sec:OccurrenceTyping}).

\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure, label=lst:numvec2]
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [x y]
  [(if x x 0) 
   (if y y 0)])
\end{lstlisting}

To check Listing \ref{lst:numvec2}, occurrence typing infers type information based on the results of conditional tests.
A conditional in Clojure is written \lstinline|(if test then else)|, where \lstinline|then|
is executed if \lstinline|test| is a true value, or \lstinline|else| is executed if \lstinline|test|
is a false value.
In Clojure, \lstinline|nil| and \lstinline|false| are the only false values: all other values are true values.
\lstinline|(if x x 0)| in Listing \ref{lst:numvec2} chooses a branch based on the truth value of the test expression \lstinline|x|.
Before this expression, the binding \lstinline|x| is known to be of type \lstinline|(U nil Number)|.
There are two cases to consider when checking this conditional: when the test \lstinline|x| is a true value, and when it is a false value.
If \lstinline|x| is a true value, then the binding \lstinline|x| must be of type \lstinline|Number|.
If \lstinline|x| is a false value, then the binding \lstinline|x| must be of type \lstinline|nil|.
Occurrence typing understands these invariants and applies them when checking each branch:
occurrences of \lstinline|x| in the first branch are considered to be of type \lstinline|Number|, and
occurrences of \lstinline|x| in the second branch are considered to be of type \lstinline|nil|.
This infers \lstinline|[(if x x 0) (if y y 0)])| to be of type
\lstinline|(Vector* Number Number)| (the type for a two-place heterogeneous vector of \lstinline|Number|s).

\subsection{Java Interoperability}
\label{sec:javainterop}

Typed Clojure supports integration with Java's type system during
interoperability from Clojure. This offers many of the same advantages
as using Java's type system from Java, for example it is a type
error to pass arguments of the wrong type to methods.
Typed Clojure also makes different decisions to Java's type system,
particularly in the treatment of Java's \emph{null}.
Note that these decisions do not change Java's type system, but only
how Typed Clojure handles Java interoperability compared to how someone
familiar with Java might expect.

From the perspective of static types in Java, \emph{null} is included in all reference types.
\emph{null} is represented in Clojure by the value \lstinline|nil|. Unlike Java's type system
Typed Clojure explicitly separates \emph{null} and reference types giving Typed Clojure 
more accurate types: we can directly express \emph{nullable} and \emph{not nullable} types.
(A type is \emph{nullable} if it may also be \emph{null},
which is expressed in Typed Clojure by creating a union of the type and \lstinline|nil|.
A type is \emph{not nullable} if it doesn't include \emph{null}).
%bump over the page
\\\\
\begin{lstlisting}[caption=Java interoperability with Typed Clojure, label=lst:interopfile]
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

(ann f File)
(def f (File. "a"))

(ann prt (U nil String))
(def prt (.getParent ^File f))

(non-nil-return java.io.File/getName :all)
(ann nme String)
(def nme (.getName ^File f))

\end{lstlisting}

Listing \ref{lst:interopfile} shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}.

Typed Clojure checks calls to Java constructors by requiring the provided
arguments be acceptable input to at least one constructor for that Class.
In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
argument, so \lstinline|(File. "a")| is type safe. Java constructors never
return \emph{null}, so Typed Clojure assigns the return type to be \lstinline|File|.
This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.

Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
By default, the \emph{java.io.File} instance method \emph{getParent}
is assigned the type \lstinline|[-> (U nil String)]| in Typed Clojure. This happens to be
a valid approximation of the method as \emph{getParent} returns \emph{null} 
``if the pathname does not name a parent directory''\footnote{See Javadoc for \emph{java.io.File}: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
On the other hand, the \emph{java.io.File} instance method \emph{getName} always returns an
instance of \emph{java.lang.String}, so we set the return position of
\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
(the second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
to override all arities of that method).

Such annotations are in a sense assumptions; should they turn out to be wrong, Typed Clojure
will infer incorrect types. 
This is somewhat troubling, but type safety can be restored by using other techniques.
For example, Typed Racket generates runtime contracts based on static types to ensure
that type errors are caught at runtime, while giving good error messages 
which indicate (or ``blame''~\cite{WF09}) the original source of the error.
A similar system for verifying that method annotations like \lstinline|non-nil-return|
are correct is planned as future work
(see section \ref{sec:contractsblame}).
